Verification of a TCP/IP stack

### 2021-05-13 SecDev21/paper.tex
Rewrite of the TCP protocol description

% Class modification
\IEEEilabelindent=0pt

@@

\section{TCP protocol}
\label{sec:TCP}

Initially described by the ``fathers of the Internet'' Vincent Cerf and Bob
Kahn, the TCP protocol was standardized in 1981 by RFC 793~\cite{rfc793}.
Its sturdiness makes it still one of the most widely used protocols today with
UDP.

In the OSI networking communication model, the TCP protocol is central and
located in the \emph{Transport layer}. It itself uses the \emph{Network layer}
to work.
The security of the \emph{Application} protocols depends on the security of
the \emph{Session layer} and consequently on the security of the TCP and UDP
protocols. Figure~\ref{Fig:TcpStack} gives an overview of how protocols are
stacked.

The purpose of this section is to give enough context to enable the reader to
follow the work conducted in this article.
The TCP protocol is connection-oriented in the sense that a connection requires
to be established between the two processes that exchange data.
The protocol is governed by a state machine described in
figure~\ref{Fig:statemachine}.
The implementation is rather free and lots of choices are left to the
programmer. However, the implementation must ensure some properties to be able
to communicate with other TCPs.

@@

\subsection{States}

@@

The behavior of the three-phase communication can be described by the state
machine.
Every group of states corresponds to a one phase. More precisely \sclosed
stands for a closed connection. Both states \ssynsent and \ssynrcv denote the
state that a TCP process can take during the initiation of a new connection. The
state \sestab represents the state when the connection is established,
\textit{i.e.} when two TCPs can communicate by sending and receiving data.
Then, two groups of states can be distinguished: the group formed by \sfwone,
\sfwtwo, \sclosing and \stimewait and that represents the state that a TCP can
reach when it is closing its connection and it is at the origin of the closing;
and the group formed by the state \sclosew and \slastack when the connection is
closing and the closing was initiated by the remote TCP.
Finally, the state \slisten stands for a TCP that is waiting for an incoming
connection.

\subsection{Flags}

The transition between the states are dictated by three things: user
actions, the reception of flags and timers. When a transition is triggered, a
response is given by sending of flags. The user actions, written in italic in
the automaton, can be an active or passive opening, or a close.
The corresponding user functions are detailed in the next paragraph.
The main flags that can be sent appear in figure~\ref{Fig:statemachine}.
They are sent alongside TCP segments, in the header.

The \syn flag is sent to synchronize two TCPs when they want to initiate a
connection. With a \ack flag, the last message received by the sender is
acknowledged. \fin is sent when a TCP wants to close the connection.
Moreover, a flag named \rst can be sent or received at any time and it forces the
connection to be reset. If so, both TCPs jump to the state \sclosed.

The state machine in figure~\ref{Fig:statemachine} does not reflect
error-conditions or any action which are not connected with the state changes.
Rather, it gives an overview of all the possible states a TCP connection could
reach over its lifetime. The transition are represented in the format
$\frac{\ \ x \ \ }{\ \ y \ \ }$ where $x$ is the event that triggers the
transition and $y$ is the response given to fulfil the transition.

\subsection{User functions}

In order to use the TCP protocol in programs, the nom defines user functions that
can be invoked by the software programmer in his code. This is the well-known
\emph{socket API}, available on almost all operating systems and known on POSIX
systems as the Berkeley sockets. This API specifies the already mentioned
\code{open} and \code{close} functions, as well as functions \code{send} and
\code{receive} to send or receive data once the connection is opened. The
behavior of each function is defined depending on the state of the TCP when the
function is called.
Even if user functions can initiate a change of state, it appears in the state
machine that most of transition are initiated by the reception of flags.

\subsection{Tasks}

A model with multiple tasks is adopted in the norm. Each task has a dedicated
job. One task is dedicated to the sending and the reception of messages. Another
is dedicated to the user functions and the last to the timers. Timers can resend
data or close a connection if the remote server does not acknowledged a message.
The first task mentioned performs most of the transitions.

@@

>>>>

@@

\section{Overview of the TCP protocol specification}
\label{sec:TCP}

This section gives an overview of the TCP functional specification without going
into unnecessary depth. A complete description of the protocol’s specification
can be found at the RFC 793 document~\cite{rfc793}.

Networking communication models define the processes of transferring information
from one network component to another. Figure~\ref{Fig:TcpStack}
shows an overview of the OSI model with its different layers.
Layering is an essential design
requirement for both models to achieve modularity, flexibility, and abstraction.
Applications that only need to use the functionality of the lower levels can
drop all the unnecessary upper levels. Furthermore, layering enables easier
troubleshooting as the network's functionality is well-defined and split between
the different layers. Each layer supports a wide range of protocols that are
compatible with the layer's specifications. The lowest layers, bottom layers in
Figure~\ref{Fig:TcpStack},
are more hardware-oriented, while the top layers are more software specific.
While the \emph{application} level
has a wide range of protocols that can be used depending on the application
specification, the \emph{transport} layer is heavily depended on the two
dominant protocols: Transmission Control Protocol (TCP) and
User Datagram Protocol (UDP).

TCP is a connection-oriented protocol; a
connection between the sender and the receiver must be established before
transmitting any data. Furthermore, TCP is a reliable protocol since it
guarantees the delivery of all the messages and that they are delivered in the
order they were sent. It also provides error-checking mechanisms to discard and
recover any corrupted data.

Initially described by the ``fathers of the Internet'' Vincent Cerf and Bob
Kahn, the TCP protocol was standardized in 1981 in~\cite{rfc793}.
Its sturdiness still makes it one of the most widely used protocols 40 years
later. Indeed, almost any networking application is using either the TCP or the
UDP protocols to traffic data between the Network and the Application layers.

@@

\subsection{TCP state machine}

@@

The behavior of the three-phase communication can be described by the state
machine in figure~\ref{Fig:statemachine}.

An edge represents a state transition. A transition is either triggered by an
explicit action, stated in italics format on the label's edge, for example,
\textit{Close}, or is triggered by the arrival of a specific flag/s.
An edge-label in the format of $\frac{\ \ x \ \ }{\ \ y \ \ }$
represents the associated flags transmitted with each transition, where $x$
is either send or receive and $y$ refers to the actual
flag/s transmitted in response to the action that caused the state-transition.
An explicit action is either a user-triggered action (through the user's API)
or an automatically performed action triggered by a timeout event. The flags are
embedded in the header section of a transmitted segment and can be one of the
following:
\begin{itemize}
\item \ack: the last message received by the sender is acknowledged.
\item \syn: this flag is sent to establish a connection.
\item \fin: this flag is sent to close the connection.
\item \rst: this flag is sent to reset the connection, often when an error
occurred in one or the other side.
\end{itemize}

The state machine in figure~\ref{Fig:statemachine} does not represent the
complete protocol specification; for example, it does not reflect
error-conditions or any actions which are not connected with the state changes.
Rather, it gives an overview of all the possible states a TCP connection could
reach over its lifetime.

Each state has a meaning. \slisten represents waiting for a connection request
from any remote TCP and port. \ssynsent and \ssynrcv represent the
initialization of the connection, that ends when \sestab is reached.
\sfwone, \sfwtwo, \sclosing and \stimewait are a group of states that represents
waiting for a connection termination request from remote TCP. \sclosew and
\slastack represent waiting for a connection termination request from local
user. Finally, \sclosed represents no connection state at all.

\subsection{TCP Multitasking model}

Different tasks can interact and update the socket data structure to handle the
various events that can occur within a TCP session. The TCP norm, under
Section “3.9. Event Processing” page 52, describes a possible implementation
of how to handle these events based on three tasks: one for the
\emph{user calls}, one for the \emph{arriving segments} and one for the
\emph{timers}.
The role of each of the three tasks can be summarized as follow:
User calls refer to functions, namely \texttt{open}, \texttt{close},
\texttt{abort}, \texttt{send} and \texttt{receive}, that can be called
by the user to control the connection, send, or receive data.
These functions can trigger transitions between the connection's states since
they are intended to control the connection.
In the arriving segments task, the received segments are being processed,
and the corresponding messages are sent back. Transitions between
states can be triggered on the reception of a segment.
Finally, timers control the timeouts, for example, the retransmission
timeout to retransmit a message, or the time-wait timeout to close the
connection after a specified amount of time elapsed. Thus, corresponding
transitions to the timeout events can also be triggered by this task.

@@

In the CycloneTCP implementation, the socket data structure is used to retain
the status of a TCP connection. Furthermore, a socket can be
manipulated by the user through the TCP's library Application Programming
Interface (API) to perform various TCP operations, such as the transmission of
data.
